---
title: First Derivative of the Multivariate Normal Densities with RcppArmadillo
author: Joscha Legewie
license: GPL (>= 2)
tags: armadillo
summary: Fast implementation of the first derivative of the Multivariate Normal density using RcppArmadillo.
---

There is a great RcppArmadillo
[implementation](http://gallery.rcpp.org/articles/dmvnorm_arma/) of
multivariate normal densities. But I was looking for the first derivative of
the multivariate normal densities. Good implementations are surprisingly hard
to come by. I wasn't able to find any online and my first solutions in R were
pretty slow. RcppArmadillo might be a great alternative particularly because
I am not aware of any c or Fortran versions that can be called from R. In
such situations, we can expect the largest performance gains. Indeed, the
RcppArmadillo version presented below is over 300-times faster than the R
implementation!

Let us start with some R code. First, `dmvnorm_deriv1` is a simple R
implementation of the formula shown in the [Matrix Cookbook](http://www2.imm.dtu.dk/pubdb/views/edoc_download.php/3274/pdf/imm3274.pdf) (formula 346 and 347,
Nov 15, 2012 version). The second version `dmvnorm_deriv2` extends Peter
Rossi's implementation of the multivariate normal densities in his package
`bayesm`, which uses a much faster algorithm and can be used to improve our
implementation of the first derivative.

```{r}
library(RcppArmadillo,quietly=TRUE)
library(rbenchmark,quietly=TRUE)
library(mvtnorm,quietly=TRUE)

dmvnorm_deriv1 <- function(X, mu=rep(0,ncol(X)), sigma=diag(ncol(X))) {
    fn <- function(x) {
        -1 * c((1/sqrt(det(2*pi*sigma))) * 
             exp(-0.5*t(x-mu)%*%solve(sigma)%*%(x-mu))) * solve(sigma,(x-mu))
    }
    out <- t(apply(X,1,fn))
    return(out)
}

# mv normal density based on Peter Rossi's implementation in `bayesm`
dMvn <- function(X,mu,Sigma) {
    k <- ncol(X)
    rooti <- backsolve(chol(Sigma),diag(k))
    quads <- colSums((crossprod(rooti,(t(X)-mu)))^2)
    return(exp(-(k/2)*log(2*pi) + sum(log(diag(rooti))) - .5*quads))
}

dmvnorm_deriv2 <- function(X, mean, sigma) {
    if (is.vector(X)) X <- matrix(X, ncol = length(X))
    if (missing(mean)) mean <- rep(0, length = ncol(X))
    if (missing(sigma)) sigma <- diag(ncol(X))
    n <- nrow(X)
    mvnorm <- dMvn(X, mu = mean, Sigma = sigma)
    deriv <- array(NA,c(n,ncol(X)))
    for (i in 1:n) {
        deriv[i,] <- -mvnorm[i] * solve(sigma,(X[i,]-mean))
    }
    return(deriv)
}
```

These implementations work but they are not very fast. `dmvnorm_deriv1` is a
one-to-one translation of the formula in pure R and more efficient algorithms
exist. `dmvnorm_deriv2` uses such an algorithm from the `bayesm` package and
is significantly faster. As shown in
[this](http://gallery.rcpp.org/articles/dmvnorm_arma/) gallery post, a
translation of this algorithm to RcppArmadillo leads to further performance
improvements. But I assume that the real bottleneck is the loop for the
calculation of the derivative. So let's adopt the calculation of the
multivariate normal from the [gallery
post](http://gallery.rcpp.org/articles/dmvnorm_arma/) and translate the loop
to RcppArmadillo. After some cleaning, we get a nice RcppArmadillo
implementation called `dmvnorm_deriv_arma`.

```{r, engine='Rcpp'}
#include <RcppArmadillo.h>

const double log2pi = std::log(2.0 * M_PI);

// [[Rcpp::depends(RcppArmadillo)]]
// [[Rcpp::export]]
arma::mat dmvnorm_deriv_arma(arma::mat x,
                             arma::rowvec mean,
                             arma::mat sigma) {

    int xdim = x.n_cols;
    arma::mat deriv;
    deriv.copy_size(x);
    arma::mat rooti = arma::trans(arma::inv(trimatu(arma::chol(sigma))));
    double rootisum = arma::sum(log(rooti.diag()));
    double constants = -(xdim/2) * log2pi;

    int n = x.n_rows;
    for (int i=0; i < n; i++) {
        arma::vec x_centered = arma::trans(x.row(i) - mean);
        arma::vec z = rooti * x_centered;
        // get derivative of multivariate normal
        deriv.row(i) = -1 * exp(constants - 
	      0.5 * arma::sum(z%z) + rootisum) * trans(solve(sigma, x_centered));
        // The part `exp(constants - 0.5 * arma::sum(z%z) + rootisum)`
	// returns the multivarate normal and the other terms translate it 
	// to the first derivative
    }

    return(deriv);
}
```

Finally, we can compare the different versions using simulated data.

```{r}
set.seed(123456789)
s <- rWishart(1, 2, diag(2))[,,1]
m <- rnorm(2)
X <- rmvnorm(10000, m, s)

benchmark(dmvnorm_deriv_arma(X,m,s),
          dmvnorm_deriv1(X,mu=m,sigma=s),
          dmvnorm_deriv2(X,mean=m,sigma=s),
          order="relative", replications=50)[,1:4]
```

The RcppArmadillo implementation is several hundred times faster! Such
stunning performance increases are possible when existing implementation rely
on pure R. Of course, the R implementation can be improved as well.
