---
title: Faster Multivariate Normal densities with RcppArmadillo and OpenMP
author: Nino Hardt, Dicko Ahmadou
license: GPL (>= 2)
tags: armadillo openmp
summary: Fast implementation of Multivariate Normal density using RcppArmadillo and OpenMP.
---

The Multivariate Normal density function is used frequently
in a number of problems. Especially for MCMC problems, fast 
evaluation is important. Multivariate Normal Likelihoods, 
Priors and mixtures of Multivariate Normals require numerous 
evaluations, thus speed of computation is vital. 
We show a twofold increase in speed by using RcppArmadillo, 
and some extra gain by using OpenMP.

This project is based on this [StackOverflow post](http://stackoverflow.com/questions/17617618/dmvnorm-mvn-density-rcpparmadillo-implementation-slower-than-r-package-includi). 

Loading necessary packages:
```{r}
libs <- c("mvtnorm","RcppArmadillo","rbenchmark","bayesm","parallel")
if (sum(!(libs %in% .packages(all.available = TRUE))) > 0) {
    install.packages(libs[!(libs %in% .packages(all.available = TRUE))])
}

for (i in 1:length(libs)) {
    library(libs[i],character.only = TRUE,quietly=TRUE)
}
```


First, we show the RcppArmadillo implementation without OpenMP.
```{r, engine='Rcpp'}
#include <RcppArmadillo.h>

// [[Rcpp::depends(RcppArmadillo)]]
// [[Rcpp::export]]
arma::vec Mahalanobis(arma::mat x, arma::rowvec center, arma::mat cov){
    int n = x.n_rows;
    arma::mat x_cen;
    x_cen.copy_size(x);
    for (int i=0; i < n; i++) {
        x_cen.row(i) = x.row(i) - center;
    }
    return sum((x_cen * cov.i()) % x_cen, 1);    
}

// [[Rcpp::export]]
arma::vec dmvnorm_arma(arma::mat x,  arma::rowvec mean,  arma::mat sigma, bool log = false) { 
    arma::vec distval = Mahalanobis(x,  mean, sigma);
    double logdet = sum(arma::log(arma::eig_sym(sigma)));
    double log2pi = std::log(2.0 * M_PI);
    arma::vec logretval = -( (x.n_cols * log2pi + logdet + distval)/2  ) ;
    
    if (log){ 
        return(logretval);
    }else { 
        return(exp(logretval));
    }
}
```

Now we simulate some data for benchmarking:
```{r}
set.seed(42)
sigma <- rwishart(10,diag(4))$IW
means <- rnorm(4)
X     <- rmvnorm(500000, means, sigma)
```

And run benchmark:
```{r}
benchmark(mvtnorm::dmvnorm(X,means,sigma), 
          dmvnorm_arma(X,means,sigma), 
 	  order="relative", replications=50)[,1:4]
```


For the OpenMP implementation, we need to enable OpenMP support.
One way of doing so is by adding the required compiler and linker
flags as follows:

```{r}
Sys.setenv("PKG_CXXFLAGS"="-fopenmp")
Sys.setenv("PKG_LIBS"="-fopenmp")
```

Forthcoming Rcpp releases will also be able to use the 'openmp' plugin which
has been added in SVN.

The source code only changes slightly. A dynamic 
schedule has been chosen for OpenMP, although a static schedule might be
faster  in some cases. This is left to further experimentation.

```{r, engine='Rcpp'}
#include <RcppArmadillo.h>
#include <omp.h>

// [[Rcpp::depends(RcppArmadillo)]]
// [[Rcpp::export]]
arma::vec Mahalanobis_mc(arma::mat x, arma::rowvec center, arma::mat cov, int cores=1){
    omp_set_num_threads( cores );
    int n = x.n_rows;
    arma::mat x_cen;
    x_cen.copy_size(x);
    #pragma omp parallel for schedule(dynamic)   
    for (int i=0; i < n; i++) {
        x_cen.row(i) = x.row(i) - center;
    }
    return sum((x_cen * cov.i()) % x_cen, 1);    
}

// [[Rcpp::export]]
arma::vec dmvnorm_arma_mc ( arma::mat x,  arma::rowvec mean,  arma::mat sigma, bool log = false, int cores=1){ 
    arma::vec distval = Mahalanobis_mc(x,  mean, sigma, cores);
    double logdet = sum(arma::log(arma::eig_sym(sigma)));
    double log2pi = std::log(2.0 * M_PI);
    arma::vec logretval = - ((x.n_cols * log2pi + logdet + distval)/2);
    
    if (log) { 
        return(logretval);
    } else { 
        return(exp(logretval));
    }
}
```


We need to set the number of cores to be used. 
```{r}
cores <- detectCores()
```

Now we are ready to benchmark again. The speed gain of 
the OpenMP version is not big, but noticable. 

```{r}
benchmark(mvtnorm::dmvnorm(X,means,sigma), 
          dmvnorm_arma(X,means,sigma) ,
          dmvnorm_arma_mc(X,means,sigma, cores), 
	  order="relative", replications=50)[,1:4]
```

The use of RcppArmadillo brings about a significant increase 
in speed. The addition of OpenMP leads to only little 
additional performance. The largest share of the increase 
in speed is due to faster computation of the Mahalanobis distance, 
which is used to compute the Multivariate Normal density.

