---
title: Using the RcppArmadillo-based Implementation of R's sample()
author: Christian Gunning and Jonathan Olmsted
license: GPL (>= 2)
updated: Dec 12, 2022
tags: rng
summary: Demonstrates the use and performance of an RcppArmadillo-based `sample()` implementation
---

### Overview and Motivation

All of R's (r\*, p\*, q\*, d\*) distribution functions are available in C++ via
the [R API] [1].  R is written in C, and the R API has no concept of a vector (at
least not in the STL sense).  Consequently, R's `sample()` function can't just be
exported via the R API, despite its importance and usefulness. The purpose of
`RcppArmadilloExtensions/sample.h` (written by Christian Gunning) is to provide
this functionaility within C++ programs.

Given sampling's central role in statistical programming, it is surprising that
no standard library implementation for C or C++ is commonly available.  There
have been repeated questions about this on the Rcpp mailing list.  StackExchange
contains an extensive discussion of this question, but there is no "canonical"
implementation.

In general, it's best to use R's tried-and-true RNG-related functions leaving
the praise (and blame) for their performance to others. R's C routines for
sampling can be found in `src/main/random.c`, with a discussion of the
associated algorithms in [Ripley 87] [2].

#### Goal 

The goal is to exactly reproduce the behavior of R's sample() function in a
templated C++/Rcpp function.  So far, we've reproduced everything but R's
implementation of the Walker Alias method (used only when sampling with
replacement using >200 non-zero weights). It uses convenience functions from
Armadillo, and thus is added to RcppArmadillo as an extension.  (The hope is
that future extensions will follow!)  All you need is this simple call, which
should work for any Rcpp Vector: `RcppArmadillo::sample(x, size, replace,
prob)`.

### Dependencies

Make sure you have a recent version of `RcppArmadillo`. The earliest adequate
release is 3.800.1 or preferably release 0.3.810.0. The usual
``install.packages("RcppArmadillo")`` command will help if 
you need to update.
You are ready to go from there:
```{r}
require(RcppArmadillo)
```

### Account for R 3.6.0 Change 

This post was originally written in 2013. Years later, with R release 3.6.0,
a change was made to how R generates random integers. This affects the
`sample()` function discussed here. To restore the behaviour present when
this post was written (and to which the C++ implementation is calibrated) use
`RNGkind(sample.kind = "Rounding").

### Quick Example

Here's a quick test to make sure it works.

Some C++ code that can be hooked into with `sourceCpp()`:

```{r engine='Rcpp'}
// [[Rcpp::depends(RcppArmadillo)]]

#include <RcppArmadilloExtensions/sample.h>

using namespace Rcpp ;

// [[Rcpp::export]]
CharacterVector csample_char( CharacterVector x, 
                              int size,
                              bool replace, 
                              NumericVector prob = NumericVector::create()
                              ) {
  CharacterVector ret = RcppArmadillo::sample(x, size, replace, prob) ;
  return ret ;
}
```

Notice that we only need `#include <RcppArmadilloExtensions/sample.h>` because
`sample.h` then `#include`-s RcppArmadillo.


We invoke the (automatically defined) `csample_char()` R function:
```{r}
N <- 10
RNGkind(sample.kind = "Rounding")  # pre R 3.6.0 behavior
set.seed(7)

sample.r <- sample(letters, N, replace=T)

set.seed(7)
sample.c <- csample_char(letters, N, replace=T)

print(identical(sample.r, sample.c))
```

Of course, R's sample() function is "internally" vectorized and
already fast.  This functionality was *not* added to speed up
`sample()`!  Instead, this lets you stay in C++ when you need to sample
from an Rcpp Vector, be it Numeric, Character, or whatever else. 

### Performance 

That said, performance is still a concern. A quick test shows a dead-heat for
sampling with replacement when compared to vanilla R:

```{r engine='Rcpp'}
#include <RcppArmadilloExtensions/sample.h>
// [[Rcpp::depends(RcppArmadillo)]]

using namespace Rcpp ;

// [[Rcpp::export]]
NumericVector csample_num( NumericVector x,
                           int size,
                           bool replace,
                           NumericVector prob = NumericVector::create()
                           ) {
  NumericVector ret = RcppArmadillo::sample(x, size, replace, prob);
  return ret;
}
```

Consider the following timing where we compare vanilla R's `sample()` to
`RcppArmadillo::sample()`. See the results for sampling with replacement with
and without probability weights.

```{r}
require(rbenchmark)
set.seed(7)

## Definition of Sampling Frame
n.elem <- 1e2
frame1 <- rnorm(n.elem)
probs1 <- runif(n.elem)

## Definition of sampling regime
## Use replacement throughout
.replace <- TRUE
## Samplesize
n.samples1 <- 1e4

## Without probabilities
benchmark(r = sample(frame1, n.samples1, replace=.replace),        
          cpp = csample_num(frame1, n.samples1, replace=.replace),
          replications = 1e3,
          order = 'relative',
          columns = c("test", "replications", "relative", "elapsed")
          )

## With probabilities
benchmark(r.prob = sample(frame1, n.samples1, prob = probs1, replace = .replace),
          cpp.prob = csample_num(frame1, n.samples1, prob = probs1, replace = .replace),
          replications = 1e3,
          order = 'relative',
          columns = c("test", "replications", "relative", "elapsed")
          )
```

The two perform equally well.

Next we look at the performance of sampling without replacement.  The number of draws can be no larger than the number of elements.  Thus we're sampling fewer elements.  Otherwise, the code is identical.

```{r}
## Use the same sampling frame as before
## Definition of sampling regime
## No replacement
.replace <- FALSE
## Since sample size can't exceed number elements, set them equal
n.samples2 <- n.elem

## Without probabilities
benchmark(r = sample(frame1, n.samples2, replace=.replace),        
          cpp = csample_num(frame1, n.samples2, replace=.replace),
          replications = 1e3,
          order = 'relative',
          columns = c("test", "replications", "relative", "elapsed")
          )

## With probabilities
benchmark(r.prob = sample(frame1, n.samples2, prob = probs1, replace = .replace),
          cpp.prob = csample_num(frame1, n.samples2, prob = probs1, replace = .replace),
          replications = 1e3,
          order = 'relative',
          columns = c("test", "replications", "relative", "elapsed")
          )
```

Finally, what we haven't done.  For sampling with replacement and more than
200 non-zero weights, R uses Walker's Alias method.  This method can be
substantially faster than the vanilla sampling method (with replacement, less
than 200 non-zero weights).  Rather than risk leading users astray with
inefficient and inappropriate methods, we throw an error.


```{r}
## Definition of Sampling Frame
n.elem <- 1e3
frame2 <- rnorm(n.elem)
probs2 <- runif(n.elem)

## Definition of sampling regime
## Use replacement throughout
.replace <- TRUE
## Samplesize
n.samples1 <- 1e4

## With probabilities
r.prob <- sample(frame2, n.samples1, prob = probs2, replace = .replace)
cpp.prob <- csample_num(frame2, n.samples1, prob = probs2, replace = .replace)
```

[1]: http://cran.r-project.org/doc/manuals/R-exts.html#Random-numbers
[2]: http://onlinelibrary.wiley.com/book/10.1002/9780470316726
