---
title: Generating a multivariate gaussian distribution using RcppArmadillo
author: Ahmadou Dicko
license: GPL (>= 2)
tags: matrix armadillo random number
summary: Demonstrate how to sample from a multivariate gaussian using a Cholesky decomposition
---

There are many ways to simulate a multivariate gaussian distribution assuming that you can simulate from independent univariate normal distributions. 
One of the most popular method is based on the [Cholesky decomposition][1].
Let's see how `Rcpp` and `Armadillo` perform on this task.


```{r engine='Rcpp'}
#include <RcppArmadillo.h>
// [[Rcpp::depends(RcppArmadillo)]]

using namespace Rcpp;

// [[Rcpp::export]]
arma::mat mvrnormArma(int n, arma::vec mu, arma::mat sigma) {
   int ncols = sigma.n_cols;
   arma::mat Y = arma::randn(n, ncols);
   return arma::repmat(mu, 1, n).t() + Y * arma::chol(sigma);
}
```


The easiest way to perform a Cholesky distribution in R is to use  the ``chol`` function in `R` which interface some fast `LAPACK` routines.


```{r}
### naive implementation in R
mvrnormR <- function(n, mu, sigma) {
    ncols <- ncol(sigma)
    mu <- rep(mu, each = n) ## not obliged to use a matrix (recycling)
    mu + matrix(rnorm(n * ncols), ncol = ncols) %*% chol(sigma)
}
```

We will also use ``MASS:mvrnorm`` which implemented it differently

```{r}
require(MASS)
### Covariance matrix and mean vector
sigma <- matrix(c(1, 0.9, -0.3, 0.9, 1, -0.4, -0.3, -0.4, 1), ncol = 3)
mu <- c(10, 5, -3)

require(MASS)
### checking variance
set.seed(123)
cor(mvrnormR(100, mu,  sigma))
cor(MASS::mvrnorm(100, mu, sigma))
cor(mvrnormArma(100, mu, sigma))

## checking means
colMeans(mvrnormR(100, mu, sigma))
colMeans(MASS::mvrnorm(100, mu, sigma))
colMeans(mvrnormArma(100, mu, sigma))
```

Now, let's benchmark the different versions

```{r}
require(rbenchmark)
benchmark(mvrnormR(1e4, mu, sigma),
          MASS::mvrnorm(1e4, mu, sigma),
          mvrnormArma(1e4, mu, sigma),
          columns = c('test', 'replications', 'relative', 'elapsed'),
          order = 'elapsed')
```

The ``RcppArmadillo`` function outperforms the MASS implementation and the naive R code, but more surprisinugly ``mvrnormR`` is slightly faster than ``mvrnorm`` in this benchmark.

To be fair, while digging into the ``MASS::mvrnorm`` code it appears that there are few code sanity checks ( such as the positive definiteness  of `Sigma` ).



[1]: http://en.wikipedia.org/wiki/Cholesky_decomposition
