---
title: Call Python from R through Rcpp
author: Wush Wu
license: GPL (>= 2)
tags: featured boost 
summary: Integrate Python into R via Rcpp and Boost.Python
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(cache=FALSE)
```

## Introduction

This post provides a brief introduction to calling Python from R through Rcpp. The
official [Python documentation](https://docs.python.org/2/extending/embedding.html) explains how to
embed python into C/C++ applications. Moreover, the
[Boost.Python](http://www.boost.org/doc/libs/1_55_0/libs/python/doc/) library
provides seamless interoperability between C++ and the Python programming
language. Similarlly, [Rcpp](http://www.rcpp.org) provides interoperability between C++ and
R. Therefore, it is not hard to call Python from R through Rcpp and Boost.Python.

Although there is a package
[*rPython*](http://cran.r-project.org/package=rPython) which provides an
interface to Python from R through Java, it is interesting to try to connect
R and Python via C++.

In this article, we show how to call Python 2.7 from R on Ubuntu. 


## Hello World

The most difficult thing is to establish a development environment. On Ubuntu,
we need to install the following packages to build via embeded Python:

```
sudo apt-get install python2.7 python2.7-dev libboost-python-dev
```

Then, we pass the following flags to the compiler:

```{r set_flags}
py_cflags <- system("python2.7-config --cflags", intern=TRUE)
Sys.setenv("PKG_CFLAGS"=sprintf("%s %s", Sys.getenv("PKG_CFLAGS"), py_cflags))
Sys.setenv("PKG_CXXFLAGS"=sprintf("%s %s", Sys.getenv("PKG_CXXFLAGS"), py_cflags))
py_ldflags <- system("python2.7-config --ldflags", intern=TRUE)
Sys.setenv("PKG_LIBS"=sprintf("%s %s %s", Sys.getenv("PKG_CFLAGS"), "-lboost_python-py27", py_ldflags))
```

The following `hello world` should then work:

```{r hello_python, engine='Rcpp', dependson='set_flags'}
#include <Rcpp.h>
#include <Python.h>

using namespace Rcpp;

//[[Rcpp::export]]
void initialize_python() {
    Py_SetProgramName("");  /* optional but recommended */
    Py_Initialize();
}

//[[Rcpp::export]]
void finalize_python() {
    Py_Finalize();
}

//[[Rcpp::export]]
void hello_python() {
    PyRun_SimpleString("from time import time,ctime\n"
                       "print 'Today is',ctime(time())\n");
}
```

```{r redirect, include=FALSE, engine='Rcpp', dependson='hello_python'}
// The following code redirect the Python stdout to R console so that these examples show the result correctly in the generated html files.
#include <Rcpp.h>
#include <Python.h>

static PyObject* redirection_stdoutredirect(PyObject* self, PyObject *args) {
    const char* string;
    if (!PyArg_ParseTuple(args, "s", &string))
       return NULL;
    Rcpp::Rcout << string;
    Py_INCREF(Py_None);
    return Py_None;
}

PyMethodDef RedirectionMethods[] = {
    {"stdoutredirect", redirection_stdoutredirect, METH_VARARGS, 
     "stdout redirection helper"},
    {NULL, NULL, 0, NULL}
};

//[[Rcpp::export]]
void Rredirect() {
    Py_InitModule("redirection", RedirectionMethods);
    PyRun_SimpleString("import sys");
    PyRun_SimpleString("import redirection");
    PyRun_SimpleString("class StdoutCatcher:\n\tdef write(self, stuff):\n\t\tredirection.stdoutredirect(stuff)\n\nsys.stdout = StdoutCatcher()\nsys.stderr = StdoutCatcher()");
}
```

```{r initialize, include=FALSE, dependson="redirect"}
initialize_python()
Rredirect()
```

Let's call them in R:

```{r hello_python_show, dependson='initialize', results='hold', eval=FALSE}
initialize_python()
hello_python()
```

```{r hello_python_run, dependson='initialize', results='hold', echo=FALSE}
hello_python()
```

It shows that the `hello_python` function successfully initializes the Python
engine and runs the Python script through `PyRun_SimpleString`.

## Type Conversion

With Boost.Python and Rcpp, we can easily transfer the data between R and
Python. The following C codes transfer the R `IntegerVector` to Python
`List`:

```{r integer_vector_transformation_cpp, engine='Rcpp', dependson='initialize'}
#include <Rcpp.h>
#include <boost/python/raw_function.hpp>

namespace py = boost::python;

typedef Rcpp::XPtr<py::list> PyList;

using namespace Rcpp;

//[[Rcpp::export]]
SEXP IntVec_to_py_list(IntegerVector src) {
    PyList pretval(new py::list());
    int glue;
    for(int i = 0;i < src.size();i++) {
        glue = src[i];
        pretval->append(glue);
    }
    return pretval;
}
```

```{r integer_vector_transformation_R, dependson="integer_vector_transformation_cpp"}
IntVec_to_py_list(1:10)
```

The pointer refers to the memory of the transformed Python object.

## Call Python Function

The following example shows how to define a function in Python and expose it in R.

```{r pycall, engine='Rcpp', dependson='initialize'}
#include <Rcpp.h>
#include <Python.h>
#include <boost/python/raw_function.hpp>

namespace py = boost::python;

typedef Rcpp::XPtr<py::list> PyList;

using namespace Rcpp;

//[[Rcpp::export]]
void pycall(std::string py_script) {
    PyRun_SimpleString(py_script.c_str());
}

//[[Rcpp::export]]
void pyfun(std::string fun_name, SEXP fun_argument) {
    // create the module of python which is similar to the R_GlobalEnv
    py::object module((py::handle<>(py::borrowed(PyImport_AddModule("__main__")))));
    // look up and retrieve the function of the given name in the module
    py::object pyfun = module.attr("__dict__")[fun_name.c_str()];
    // call the function with the API of boost::python
    py::list argv(*PyList(fun_argument));
    pyfun(argv);
}
```

```{r use_list, dependson='py'}
pycall("
def print_list(src):
    for i in src:
        print i
")
a <- IntVec_to_py_list(1:10)
pyfun("print_list", a)
```

## Error Handling

The error of Python engine could be handled easily by the C++ try/catch as follow:

```{r py_exception, engine='Rcpp', dependson='pycall'}
#include <Rcpp.h>
#include <Python.h>
#include <boost/python/raw_function.hpp>

namespace py = boost::python;

typedef Rcpp::XPtr<py::list> PyList;

//[[Rcpp::export]]
void pyfun(std::string fun_name, SEXP fun_argument) {
    try {
        // create the module of python which is similar to the R_GlobalEnv
        py::object module((py::handle<>(py::borrowed(PyImport_AddModule("__main__")))));
        // look up and retrieve the function of the given name in the module
        py::object pyfun = module.attr("__dict__")[fun_name.c_str()];
        // call the function with the API of boost::python
        py::list argv(*PyList(fun_argument));
        pyfun(argv);
    }
    catch (py::error_already_set) {
        PyErr_Print();
    }
}
```

```{r py_exception_demo, dependson='py_exception'}
pycall("
def print_list(src):
    for i in src:
        print i
")
a <- IntVec_to_py_list(1:10)
pyfun("print_lists", a) # a typo of the function name
```

## Summary

These examples show how to integrate Python and R with Rcpp and
Boost.Python. Similar approaches could be used to integrate R with other
scripting language such as javascript(nodejs). The core steps are
initializing the engine (Hello World), transforming the data (Type
Conversion), exposing the functions (Call Python Function), and handling the
error properly (Error Handling). Some C++ libraries ease the work greatly such
as Rcpp for R, Boost.Python for Python and
[V8Converter](https://code.google.com/p/v8-juice/) for javascript.
