---
title: Create R-tree data structure using Rcpp and BH package(Boost::Geometry).
author: teramonagi
license: GPL (>= 2)
mathjax: false
tags: boost 
summary: In this document, We explain the basic concept of Boost::Geometry and R-tree, and show how to create a wrapper class of rtree class in Boost::Geometry library using Rcpp. 
---

## Introduction

The purpose of this gallery post is to show you how to use [Boost::Geometry library](http://www.boost.org/doc/libs/1_57_0/libs/geometry/doc/html/index.html) which was introduced [recently](http://dirk.eddelbuettel.com/blog/2014/12/22/#bh_1.55.0-1) in Rcpp.
Especially, we focus on [R-tree data structure](http://www.boost.org/doc/libs/1_57_0/libs/geometry/doc/html/geometry/reference/spatial_indexes/boost__geometry__index__rtree.html) for searching objects in space because only one spatial index is implemented - R-tree Currently in this library.

Boost.Geometry which is part of the Boost C++ Libraries gives us algorithms for solving geometry problems. In this library, the Boost.Geometry.Index which is one of components is intended to gather data structures called spatial indexes which are often used to searching for objects in space quickly. Generally speaking, spatial indexes stores *geometric objects' representations* and allows searching for objects occupying some space or close to some point in space.

## R-tree

R-tree is a tree data structure used for spatial searching, i.e., for indexing multi-dimensional information such as geographical coordinates, rectangles or polygons. R-tree was proposed by Antonin Guttman in 1984 as an expansion of B-tree for multi-dimensional data and plays significant role in both theoretical and applied contexts. It is only one spatial index implemented in Boost::Geometry.

As a real application of this, It is often used to store spatial objects such as restaurant locations or the polygons which typical maps are made of: streets, buildings, outlines of lakes, coastlines, etc in order to perform a spatial query like "Find all stations within 1 km of my current location", "Let me know all road segments in 2 km of my location" or "find the nearest gas station" which we often ask google seach by your voice recenlty. In this way, the R-tree can  be used (nearest neighbor) search for some places.

You can find more explanations about R-tree in [Wikipedia](http://en.wikipedia.org/wiki/R-tree).

## Write a wrapper class of rtree in Boost::Geometry using Rcpp

Now, we write a simple C++ wrapper class of rtree class in Boost::Geometry::Index that we can use  in R.

The most important feature to mention here is the use of Rcpp module to expose 
your own class to R. Although almost all classes in Boost library have a lot of functions,
, you do not use all in many cases. In that case, you should write your wrapper class for making your code simple.

### Rcpp code

```{r, engine = "Rcpp"}
// [[Rcpp::depends(BH)]]
#include <vector>
#include <Rcpp.h>
#include <boost/geometry.hpp>
#include <boost/geometry/geometries/point.hpp>
#include <boost/geometry/geometries/box.hpp>
#include <boost/geometry/index/rtree.hpp>
using namespace Rcpp;

//Mnemonics
namespace bg = boost::geometry;
namespace bgi = boost::geometry::index;
typedef bg::model::point<float, 2, bg::cs::cartesian> point_t;
typedef bg::model::box<point_t> box;
typedef std::pair<box, unsigned int> value_t;

//Forward declaration for RTreeCpp class and export setting.
class RTreeCpp;
RCPP_EXPOSED_CLASS(RTreeCpp)

class RTreeCpp
{
public:
  //Constructor.
  //You have to give spatial data as a data frame.
  RTreeCpp(DataFrame df)
  {
      int size = df.nrows();
      NumericVector id   = df[0]; 
      NumericVector bl_x = df[1]; 
      NumericVector bl_y = df[2]; 
      NumericVector ur_x = df[3]; 
      NumericVector ur_y = df[4]; 
      for(int i = 0; i < size; ++i)
      {        
        // create a box
        box b(point_t(bl_x[i], bl_y[i]), point_t(ur_x[i], ur_y[i]));
        // insert new value
        rtree_.insert(std::make_pair(b, static_cast<unsigned int>(id[i])));   
      }
  }
  //This method(query) is k-nearest neighbor search. 
  //It returns some number of values nearest to some point(point argument) in space.
  IntegerVector knn(NumericVector point, unsigned int n)
  {
    std::vector<value_t> result_n;
    rtree_.query(bgi::nearest(point_t(point[0], point[1]), n), std::back_inserter(result_n));    
    std::vector<int> indexes;
    std::vector<value_t>::iterator itr;
    for ( itr = result_n.begin(); itr != result_n.end(); ++itr ) {
      value_t value = *itr;
      indexes.push_back( value.second );
    }
    return as<IntegerVector>(wrap(indexes)) ;
  }
private:
  //R-tree can be created using various algorithm and parameters
  //You can change the algorithm by template parameter. 
  //In this example we use quadratic algorithm. 
  //Maximum number of elements in nodes in R-tree is set to 16.
  bgi::rtree<value_t, bgi::quadratic<16> > rtree_;
};

//Rcpp modules allow programmers to expose C++ functions/classes to R easily
RCPP_MODULE(ModTree){
    class_<RTreeCpp>("RTreeCpp")
        .constructor<Rcpp::DataFrame>()
        .method( "knn", &RTreeCpp::knn )
    ;
}
```

### R code using RTreeCpp
At first, we make a sample spatial data.
```{r}
#Sample spatial data(boxes)
points <- data.frame(
  id=0:2, 
  bl_x=c(0, 2, 4), 
  bl_y=c(0, 2, 4), 
  ur_x=c(1, 3, 5), 
  ur_y=c(1, 3, 5))
```
To show this visually, we write the following code: 
```{r}
size <- nrow(points)
#colors for rectangle area
colors <- rainbow(size)
#Plot these points
plot(c(0, 5), c(0, 5), type= "n", xlab="", ylab="")
for(i in 1:size){
  rect(points[i, "bl_x"], points[i, "bl_y"], points[i, "ur_x"], points[i, "ur_y"], col=colors[i])  
}
legend(4, 2, legend=points$id, fill=colors)
```

You can use RTreeCpp class as the following:
```{r}
#new RTreeCpp object
rt <- new(RTreeCpp, points)
#Search nearest neighbor points(return value : id of points data)
rt$knn(c(0, 0), 1)
rt$knn(c(0, 0), 2)
rt$knn(c(0, 0), 3)
```