---
title: Serialize and Deserialize a C++ Object in Rcpp
author: Wush Wu
license: GPL (>= 2)
tags: serialization
summary: We demonstrate how to serialize a c++ object to the raw vector in R and deserialize it.
---

This post shows how to serialize a c++ object to the raw vector in R and deserialize it with the help of Rcereal and BH.

First, please install the Rcpp, Rcereal, and BH from CRAN and enable the support of C++11 via `Sys.setenv("PKG_CXXFLAGS"="-std=c++11")`.

```{r c++11, include = FALSE}
Sys.setenv("PKG_CXXFLAGS"="-std=c++11")
```

We can use the cereal library and boost iostreams in Rcpp. The following example shows a toy C++ class and the related serialize/deserialize functions.

```{r, engine = "Rcpp"}
// [[Rcpp::depends(Rcereal)]]
// [[Rcpp::depends(BH)]]

#include <boost/iostreams/stream.hpp>
#include <boost/iostreams/device/array.hpp>
#include <cereal/archives/binary.hpp>
#include <Rcpp.h>

struct MyClass
{
  int x, y, z;

  // This method lets cereal know which data members to serialize
  template<class Archive>
  void serialize(Archive & archive)
  {
    archive( x, y, z ); // serialize things by passing them to the archive
  }
};

using namespace Rcpp;

//[[Rcpp::export]]
RawVector serialize_myclass(int x = 1, int y = 2, int z = 3) {
  MyClass my_instance;
  my_instance.x = x;
  my_instance.y = y;
  my_instance.z = z;
  RawVector retval(100);
  boost::iostreams::stream_buffer<boost::iostreams::array_sink> buf((char*) &retval[0], retval.size());
  std::ostream ss(&buf);
  {
    cereal::BinaryOutputArchive oarchive(ss);
    oarchive(my_instance);
  }
  return retval;
}

//[[Rcpp::export]]
void deserialize_myclass(RawVector src) {
  boost::iostreams::stream<boost::iostreams::array_source> ss((char*) &src[0], src.size());
  MyClass my_instance;
  {
    cereal::BinaryInputArchive iarchive(ss);
    iarchive(my_instance);
  }
  Rcout << my_instance.x << "," << my_instance.y << "," << my_instance.z << std::endl;
}
```

Thanks to Rcpp, the compiler will automatically find the header file according to `// [[Rcpp::depends(Rcereal)]]` and `// [[Rcpp::depends(BH)]]`.
The member function `void serialize(Archive & archive)` tells cereal how to serialize and deserialize the C++ class `MyClass`. 
Data will be saved and loaded if they are passed to the argument `archive`.

The `cereal::BinaryOutputArchive oarchive(ss);` defines an instance of bineary archive. The `oarchive(my_instance);` archives the instance of `MyClass` to the output stream and writes the data to the `RawVector`.

Similarly, `cereal::BinaryInputArchive iarchive(ss);` defines an instace of bineary archive. The `iarchive(my_instance);` reads the data from the `RawVector` and restores the content to the instance of `MyClass`.

Let's try these two functions in R:

```{r}
v <- serialize_myclass(1, 2, 4)
head(v)
deserialize_myclass(v)
```

As you can see, the instance of `MyClass` is successfully saved to `v` and loaded from `v`.

The API of cereal is similar to ones of boost serialization. The main difference is that the cereal prefers using `()` to send data to archives.
Please visit <http://uscilab.github.io/cereal/transition_from_boost.html> for more details about the difference between cereal and boost serialization.

In my opinion, the most important advantage is that cereal is header only, so it is more easily to write portable package with cereal compared to boost serialization. 

