---
title: Serialize and Deserialize a C++ Object in Rcpp
author: Wush Wu
license: GPL (>= 2)
tags: serialization
summary: We demonstrate how to serialize a C++ object into an R raw vector objects, and how to deserialize it.
---

This post shows how to serialize a C++ object into a R raw vector object---the base type used by the internal R serialization---and how to deserialize it. 

The following example shows a toy C++ class and the related serialize/deserialize functions.  It shows how one can use the
[cereal header-only C++ library](https://github.com/USCiLab/cereal) and [Boost Iostreams](http://www.boost.org/doc/libs/1_59_0/libs/iostreams/doc/index.html) in
Rcpp.  It relies on the corresponding [Rcereal](https://cran.rstudio.com/package=Rcereal) and [BH](https://cran.rstudio.com/package=BH) packages which need to be
installed.  It also requires compilation with support for the C++11 standard, which we enable via the corresponding Rcpp plugin.


```{r, engine = "Rcpp"}
// Enable C++11 via this plugin 
// [[Rcpp::plugins("cpp11")]]

// [[Rcpp::depends(Rcereal)]]
// [[Rcpp::depends(BH)]]

#include <boost/iostreams/stream.hpp>
#include <boost/iostreams/device/array.hpp>
#include <cereal/archives/binary.hpp>
#include <Rcpp.h>

struct MyClass {
    int x, y, z;

    // This method lets cereal know which data members to serialize
    template<class Archive>
    void serialize(Archive & archive) {
        archive( x, y, z ); // serialize things by passing them to the archive
    }
};

using namespace Rcpp;

//[[Rcpp::export]]
RawVector serialize_myclass(int x = 1, int y = 2, int z = 3) {
    MyClass my_instance;
    my_instance.x = x;
    my_instance.y = y;
    my_instance.z = z;
    RawVector retval(100);
    boost::iostreams::stream_buffer<boost::iostreams::array_sink> buf((char*) &retval[0], retval.size());
    std::ostream ss(&buf); {
        cereal::BinaryOutputArchive oarchive(ss);
        oarchive(my_instance);
    }
    return retval;
}

//[[Rcpp::export]]
void deserialize_myclass(RawVector src) {
    boost::iostreams::stream<boost::iostreams::array_source> ss((char*) &src[0], src.size());
    MyClass my_instance; {
        cereal::BinaryInputArchive iarchive(ss);
        iarchive(my_instance);
    }
    Rcout << my_instance.x << "," << my_instance.y << "," << my_instance.z << std::endl;
}
```

Thanks to how Rcpp and R interact, the compiler will automatically find the header file according to `// [[Rcpp::depends(Rcereal)]]` and `//
[[Rcpp::depends(BH)]]`.  The member function `void serialize(Archive & archive)` tells cereal how to serialize and deserialize the C++ class `MyClass`.  Data
will be saved and loaded if they are passed to the argument `archive`.

The `cereal::BinaryOutputArchive oarchive(ss);` defines an instance of bineary archive. The `oarchive(my_instance);` archives the instance of `MyClass` to the
output stream and writes the data to the `RawVector`.

Similarly, `cereal::BinaryInputArchive iarchive(ss);` defines an instace of bineary archive. The `iarchive(my_instance);` reads the data from the `RawVector`
and restores the content to the instance of `MyClass`.

Let's try these two functions in R:

```{r}
v <- serialize_myclass(1, 2, 4)
head(v)
deserialize_myclass(v)
```

As you can see, the instance of `MyClass` is successfully saved to `v` and loaded from `v`.

The API of cereal is similar to the one of Boost Serialization. The main difference is that the cereal prefers using `()` to send data to archives.  Please
visit the [transition from Boost](http://uscilab.github.io/cereal/transition_from_boost.html) page for more details about the difference between cereal and
Boost Serialization.

The most important advantage may be that cereal is header-only, which makes it easier to write portable package with cereal compared to Boost Serialization.

