---
title: Custom progress bars for RcppProgress
author: Karl Forner and Clemens Schmid
license: GPL (>= 2)
tags: progress bar
summary: Demonstrates how to implement custom progress bars for long computations in C++.
---

[RcppProgress](http://cran.r-project.org/web/packages/RcppProgress/index.html) 
is a tool to help you monitor the execution time of your C++ code, by
providing a way to interrupt the execution inside the C++ code, and also to
display a progress bar indicative of the state of your computation. Additionally, 
it is compatible with multithreaded code, for example using OpenMP. 
[This article](http://gallery.rcpp.org/articles/using-rcppprogress/) explains the 
basic setup.

Since version 0.4 it became more simple to create custom progress bars. We will 
showcase an example implementation here by setting up a progress bar that also
displays an estimate of the remaining time. You can find a complete package setup 
with this code [here](https://github.com/kforner/rcpp_progress/tree/master/inst/examples/RcppProgressETA). 

### ETAProgressBar

Imagine you added a progress bar with RcppProgress to your function `long_computation()` 
following the example from the first article mentioned above.
 
```{r long_computation, engine="Rcpp"}
// [[Rcpp::depends(RcppProgress)]]
#include <progress.hpp>
#include <progress_bar.hpp>
// [[Rcpp::export]]
double long_computation(int nb, bool display_progress=true) {
    double sum = 0;
    Progress p(nb*nb, display_progress);
    for (int i = 0; i < nb; ++i) {
        if (Progress::check_abort() )
            return -1.0;
        for (int j = 0; j < nb; ++j) {
            p.increment(); 
            sum += R::dlnorm(i+j, 0.0, 1.0, 0);
        }
    }
    return sum + nb;
}
```

```{r long_computation_call}
long_computation(1000)
```

You can now create your own implementation of a progress bar.

```{r eta, engine="Rcpp"}
#ifndef _RcppProgress_ETA_PROGRESS_BAR_HPP
#define _RcppProgress_ETA_PROGRESS_BAR_HPP

#include <R_ext/Print.h>
#include <ctime>
#include <stdio.h>
#include <sstream>
#include <string.h>

// [[Rcpp::depends(RcppProgress)]]
#include "progress_bar.hpp"

// for unices only
#if !defined(WIN32) && !defined(__WIN32) && !defined(__WIN32__)
#include <Rinterface.h>
#endif

class ETAProgressBar: public ProgressBar{
  public: // ====== LIFECYCLE =====
    
    /**
    * Main constructor
    */
    ETAProgressBar()  {
      _max_ticks = 50;
      _ticks_displayed = 0;
      _finalized = false;
      _mode_flag = 1;
      _timer_flag = true;
    }
    
    ~ETAProgressBar() {
    }
    
    public: // ===== main methods =====
      
      void display() {
        REprintf("0%%   10   20   30   40   50   60   70   80   90   100%%\n");
        REprintf("[----|----|----|----|----|----|----|----|----|----|\n");
        flush_console();
      }
      
      // update display
      void update(float progress) {
        
        // stop if already finalized
        if (_finalized) return;
        
        // start time measurement when update() is called the first time
        if (_timer_flag) {
          _timer_flag = false;
          // measure start time
          time(&start);
        } else {
          
          // measure current time
          time(&end);
          
          // calculate passed time and remaining time (in seconds)
          double pas_time = std::difftime(end, start);
          double rem_time = (pas_time / progress) * (1 - progress);
          
          // convert seconds to time string
          std::string time_string = _time_to_string(rem_time);
          
          // create progress bar string 
          std::string progress_bar_string = _current_ticks_display(progress);
          
          // ensure overwriting of old time info
          int empty_length = time_string.length();
          std::string empty_space = std::string(empty_length, ' ');
          
          // merge progress bar and time string
          std::stringstream strs;
          strs << "|" << progress_bar_string << "| " << time_string << empty_space;
          std::string temp_str = strs.str();
          char const* char_type = temp_str.c_str();
          
          // print: remove old and replace with new
          REprintf("\r");
          REprintf("%s", char_type);
          
          // finalize display when ready
          if(progress == 1) {
            _finalize_display();
          }  
        }
      }
      
      void end_display() {
        update(1);
      }
      
      
      protected: // ==== other instance methods =====
        
        // convert double with seconds to time string
        std::string _time_to_string(double seconds) {
          
          int time = (int) seconds;
          
          int hour = 0;
          int min = 0;
          int sec = 0;
          
          hour = time / 3600;
          time = time % 3600;
          min = time / 60;
          time = time % 60;
          sec = time;
          
          std::stringstream time_strs;
          if (hour != 0) time_strs << hour << "h ";
          if (min != 0) time_strs << min << "min ";
          if (sec != 0) time_strs << sec << "s ";
          std::string time_str = time_strs.str();
          
          return time_str;
        }
        
        // update the ticks display corresponding to progress
        std::string _current_ticks_display(float progress) {
          
          int nb_ticks = _compute_nb_ticks(progress);
          
          std::string cur_display = _construct_ticks_display_string(nb_ticks);
          
          return cur_display;
        }
        
        // construct progress bar display
        std::string _construct_ticks_display_string(int nb) {
          
          std::stringstream ticks_strs;
          for (int i = 0; i < (_max_ticks - 1); ++i) {
            if (i < nb) {
              ticks_strs << "*";
            } else {
              ticks_strs << " ";
            }
          }
          std::string tick_space_string = ticks_strs.str();
          
          return tick_space_string;
        }
        
        // finalize
        void _finalize_display() {
          if (_finalized) return;
          
          REprintf("\n");
          flush_console();
          _finalized = true;
        }
        
        // compute number of ticks according to progress
        int _compute_nb_ticks(float progress) {
          return int(progress * _max_ticks);
        }
        
        // N.B: does nothing on windows
        void flush_console() {
#if !defined(WIN32) && !defined(__WIN32) && !defined(__WIN32__)
          R_FlushConsole();
#endif
        }
        
        private: // ===== INSTANCE VARIABLES ====
          int _max_ticks;   		// the total number of ticks to print
          int _ticks_displayed; 	// the nb of ticks already displayed
          bool _finalized;
          int _mode_flag;
          bool _timer_flag;
          time_t start,end;
          
};

#endif



// [[Rcpp::depends(RcppProgress)]]
#include <progress.hpp>
// [[Rcpp::export]]
double long_computation_eta(int nb, bool display_progress=true) {
    ETAProgressBar pb;
    double sum = 0;
    Progress p(nb*nb, display_progress, pb);
    for (int i = 0; i < nb; ++i) {
        if (Progress::check_abort() )
            return -1.0;
        for (int j = 0; j < nb; ++j) {
            p.increment(); 
            sum += R::dlnorm(i+j, 0.0, 1.0, 0);
        }
    }
    return sum + nb;
}
```

```{r long_computation_2_call}
long_computation_eta(1000)
```
