---
title: "Cleaner Generic Function with RCPP_RETURN Macros"
author: "Nathan Russell"
license: GPL (>= 2)
tags: c++98 c++11 macros
summary: "This post demonstrates usage of the RCPP_RETURN macros."
---

### TL;DR 
> * C++ templates and function overloading are incompatible with R's C API, so 
>   polymorphism must be achieved via run-time dispatch, handled explicitly by 
>   the programmer.
> * The traditional technique for operating on `SEXP` objects in a generic 
>   manner entails a great deal of boilerplate code, which can be unsightly, 
>   unmaintainable, and error-prone.
> * The desire to provide polymorphic functions which operate on vectors 
>   and matrices is common enough that Rcpp provides the utility macros 
>   `RCPP_RETURN_VECTOR` and `RCPP_RETURN_MATRIX` to simplify the process. 
> * Subsequently, these macros were extended to handle an (essentially) 
>   arbitrary number of arguments, provided that a C++11 compiler is used. 

---

### Background
To motivate a discussion of polymorphic functions, imagine that we desire a 
function (`ends`) which, given an input vector `x` and an integer `n`, returns 
a vector containing the first and last `n` elements of `x` concatenated. 
Furthermore, we require `ends` to be a single interface which is capable of 
handling multiple types of input vectors (integers, floating point values, 
strings, etc.), rather than having a separate function for each case. How can 
this be achieved? 

#### R Implementation 
A naïve implementation in R might look something like this: 

```{r}
ends <- function(x, n = 6L) 
{
    n <- min(n, length(x) %/% 2)
    c(head(x, n), tail(x, n))
}

ends(1:9)

ends(letters, 3)

ends(rnorm(20), 2)
```

The simple function above demonstates a key feature of many dynamically-typed 
programming languages, one which has undoubtably been a significant factor in their 
rise to popularity: the ability to write generic code with little-to-no 
additional effort on the part of the developer. Without getting into a discussion 
of the pros and cons of static vs. dynamic typing, it is evident that being able 
to dispatch a single function generically on multiple object types, as opposed to, 
e.g. having to manage separate impementations of `ends` for each vector type, 
helps us to write more concise, expressive code. Being an article about Rcpp, 
however, the story does not end here, and we consider how this problem might 
be approached in C++, which has a *much* more strict type system than R. 

#### C++ Implementation(s)
For simplicity, we begin by considering solutions in the context of a "pure" 
(re: not called from R) C++ program. Eschewing more complicated tactics 
involving run-time dispatch (`virtual` functions, etc.), the C++ language 
provides us with two straightforward methods of achieving this at compile time: 

1. Function Overloading ([ad hoc polymorphism](https://en.wikipedia.org/wiki/Ad_hoc_polymorphism)) 
2. Templates ([parametric polymorphism](https://en.wikipedia.org/wiki/Parametric_polymorphism))

The first case can be demonstrated as follows: 

```{r, engine='Rcpp', eval=FALSE}
#include <iostream>
#include <vector>
#include <string>
#include <algorithm>

typedef std::vector<int> ivec;

ivec ends(const ivec& x, std::size_t n = 6) 
{
    n = std::min(n, x.size() / 2);
    ivec res(2 * n);

    std::copy(x.begin(), x.begin() + n, res.begin());
    std::copy(x.end() - n, x.end(), res.begin() + n);

    return res;
}

typedef std::vector<double> dvec;

dvec ends(const dvec& x, std::size_t n = 6) 
{
    n = std::min(n, x.size() / 2);
    dvec res(2 * n);

    std::copy(x.begin(), x.begin() + n, res.begin());
    std::copy(x.end() - n, x.end(), res.begin() + n);

    return res;
}

typedef std::vector<std::string> svec;
// and so on...

int main()
{
    ivec x, xres;
    dvec y, yres;

    for (int i = 0; i < 20; i++) {
        x.push_back(i);
        y.push_back(i + 0.5);
    }

    xres = ends(x, 4);
    yres = ends(y);

    for (std::size_t i = 0; i < xres.size(); i++) {
        std::cout << xres[i] << "\n";
    }

    for (std::size_t i = 0; i < yres.size(); i++) {
        std::cout << yres[i] << "\n";
    }
}
```

Although the above program meets our criteria, the code duplication is profound. 
Being seasoned C++ programmers, we recognize this 
as a textbook use case for templates and refactor accordingly: 

```{r, engine='Rcpp', eval=FALSE}
#include <iostream>
#include <vector>
#include <string>
#include <algorithm>

template <typename T>
T ends(const T& x, std::size_t n = 6) 
{
    n = std::min(n, x.size() / 2);
    T res(2 * n);

    std::copy(x.begin(), x.begin() + n, res.begin());
    std::copy(x.end() - n, x.end(), res.begin() + n);

    return res;
}

typedef std::vector<int> ivec;
typedef std::vector<double> dvec;
typedef std::vector<std::string> svec;
// and so on...

int main()
{
    // as before
}
```

This approach is much more maintainable as we have a single implementation 
of `ends` rather than one implementation per `typedef`. With this in hand, we 
now look to make our C++ version of `ends` callable from R via Rcpp. 

#### Rcpp Implementation (First Attempt)

Many people, myself included, have attempted some variation of the following at 
one point or another: 

```{r, engine='Rcpp', eval=FALSE}
#include <Rcpp.h>

// [[Rcpp::export]]
template <typename T>
T ends(const T& x, std::size_t n = 6) 
{
    n = std::min(n, x.size() / 2);
    T res(2 * n);

    std::copy(x.begin(), x.begin() + n, res.begin());
    std::copy(x.end() - n, x.end(), res.begin() + n);

    return res;
}
```

Sadly this does not work: magical as Rcpp attributes may be, there are limits 
to what they can do, and at least for the time being, translating C++ template 
functions into something compatible with R's C API is out of the question. Similarly, 
the first C++ approach from earlier is also not viable, as the C programming 
language does not support function overloading. In fact, C does not 
support *any* flavor of type-safe static polymorphism, meaning that our generic 
function must be implemented through run-time polymorphism, as touched on in 
Kevin Ushey's Gallery article [Dynamic Wrapping and Recursion with Rcpp](
http://gallery.rcpp.org/articles/rcpp-wrap-and-recurse/).

#### Rcpp Implementation (Second Attempt)

Armed with the almighty `TYPEOF` macro and a [SEXPTYPE cheatsheat](
https://cran.r-project.org/doc/manuals/r-release/R-ints.html#SEXPTYPEs), we 
modify the template code like so: 

```{r, engine='Rcpp'}
#include <Rcpp.h>
using namespace Rcpp;

namespace impl {

template <int RTYPE>
Vector<RTYPE> ends(const Vector<RTYPE>& x, int n)
{
    n = std::min((R_xlen_t)n, x.size() / 2);
    Vector<RTYPE> res(2 * n);

    std::copy(x.begin(), x.begin() + n, res.begin());
    std::copy(x.end() - n, x.end(), res.begin() + n);

    return res;
}

} // impl

// [[Rcpp::export]]
SEXP ends(SEXP x, int n = 6) {
    switch (TYPEOF(x)) {
        case INTSXP: {
            return impl::ends(as<IntegerVector>(x), n);
        }
        case REALSXP: {
            return impl::ends(as<NumericVector>(x), n);
        }
        case STRSXP: {
            return impl::ends(as<CharacterVector>(x), n);
        }
        case LGLSXP: {
            return impl::ends(as<LogicalVector>(x), n);
        }
        case CPLXSXP: {
            return impl::ends(as<ComplexVector>(x), n);
        }
        default: {
            warning(
                "Invalid SEXPTYPE %d (%s).\n",
                TYPEOF(x), type2name(x)
            );
            return R_NilValue;
        }
    }
}
```

```{r}
ends(1:9)

ends(letters, 3)

ends(rnorm(20), 2)

ends(list())
```

Some key remarks: 

1. Following the ubiquitous Rcpp idiom, we have converted our `ends` template to use 
an integer parameter instead of a type parameter. This is a crucial point, and 
later on, we will exploit it to our benefit. 
2. The template implementation is wrapped in a namespace in order to avoid a 
naming conflict; this is a personal preference but not strictly necessary. 
Alternatively, we could get rid of the namespace and rename either the template 
function or the exported function (or both). 
3. We use the opaque type `SEXP` for our input / output vector since we need a 
single input / output type. In this particular situation, replacing `SEXP` with 
the Rcpp type `RObject` would also be suitable as it is a generic class capable 
of representing any `SEXP` type. 
4. Since we have used an opaque type for our input vector, we must cast it 
to the appropriate `Rcpp::Vector` type accordingly within each case label. (For 
further reference, the list of vector aliases can be found [here](
https://github.com/RcppCore/Rcpp/blob/master/inst/include/Rcpp/vector/instantiation.h#L27-L50)). Finally, we *could* dress each return value in `Rcpp::wrap` to convert 
the `Rcpp::Vector` to a `SEXP`, but it isn't necessary because Rcpp attributes 
will do this automatically (if possible). 

At this point we have a polymorphic function, written in C++, and callable from 
R. But that `switch` statement sure is an eyesore, and it will need to be 
implemented every time we wish to export a generic function to R. Aesthetics 
aside, a more pressing concern is that boilerplate such as this increases the 
likelihood of introducing bugs into our codebase -- and since we are leveraging 
run-time dispatch, these bugs will not be caught by the compiler. For example, 
there is nothing to prevent this from compiling: 

```c++
// ...
case INTSXP: {
    return impl::ends(as<CharacterVector>(x), n);
} 
// ...
```

In our particular case, such mistakes likely would not be too disastrous, but 
it should not be difficult to see how situations like this can put you (or a 
user of your library!) on the fast track to segfault. 

---

### Obligatory Remark on Macro Safety

The C preprocessor is undeniably one of the more controversial aspects of the 
C++ programming language, as its utility as a metaprogramming tool is rivaled 
only by its potential for abuse. A proper discussion of the various pitfalls 
associated with C-style macros is well beyond the scope of this article, so 
the reader is encouraged explore this topic on their own. On the bright side, 
the particular macros that we will be discussing are sufficiently complex 
and limited in scope that misuse is much more likely to result in a compiler 
error than a silent bug, so practically speaking, one can expect a fair bit of 
return for relatively little risk.

---

### Synopsis 

At a high level, we summarize the `RCPP_RETURN` macros as follows: 

* There are two separate macros for dealing with vectors and matrices, 
  `RCPP_RETURN_VECTOR` and `RCPP_RETURN_MATRIX`, respectively. 
* In either case, code is generated for the following `SEXPTYPE`s:
    * `INTSXP` (integers)
    * `REALSXP` (numerics)
    * `RAWSXP` (raw bits)
    * `LGLSXP` (logicals)
    * `CPLXSXP` (complex numbers)
    * `STRSXP` (characters / strings)
    * `VECSXP` (lists)
    * `EXPRSXP` (expressions) 

* In C++98 mode, each macro accepts two arguments:
    1. A template function
    2. A `SEXP` object 

* In C++11 mode (or higher), each macro additionally accepts zero or more 
  arguments which are forwarded to the template function.

Finally, the template function must meet the following criteria: 

* It is templated on a single, integer parameter.
* In the C++98 case, it accepts a single `SEXP` (or something convertible to 
  `SEXP`) argument. 
* In the C++11 case, it may accept more than one argument, but the first 
  argument is subject to the previous constraint. 

Examining our templated `impl::ends` function from the previous section, we see 
that it meets the first requirement, but fails the second, due to its second 
parameter `n`. Before exploring how `ends` might be adapted to meet the (C++98) 
template requirements, it will be helpful demonstrate correct usage with a few 
simple examples. 

---

### Fixed Return Type

We consider two situations where our input type is generic, but our output 
type is fixed:

1. Determining the length (number of elements) of 
  a vector, in which an `int` is always returned.
2. Determining the dimensions (number of rows and number of columns) 
  of a matrix, in which a length-two `IntegerVector` is always returned. 
  
First, our `len` function: 

```{r, engine='Rcpp'}
#include <Rcpp.h>
using namespace Rcpp;

namespace impl {

template <int RTYPE>
int len(const Vector<RTYPE>& x) 
{
    return static_cast<int>(x.size());
}

} // impl

// [[Rcpp::export]]
int len(RObject x) 
{
    RCPP_RETURN_VECTOR(impl::len, x);
}
```

(Note that we omit the `return` keyword, as it is part of the macro definition.) 
Testing this out on the various supported vector types: 

```{r}
classes <- c(
    "integer", "numeric", "raw", "logical",
    "complex", "character", "list", "expression"
)
sapply(seq_along(classes), function(i) {
    x <- vector(mode = classes[i], length = i)
    all.equal(len(x), length(x))
})
```

Similarly, creating a generic function that determines the dimensions of an 
input matrix is trivial: 

```{r, engine='Rcpp'}
#include <Rcpp.h>
using namespace Rcpp;

namespace impl {

template <int RTYPE>
Vector<INTSXP> dims(const Matrix<RTYPE>& x) 
{
    return Vector<INTSXP>::create(x.nrow(), x.ncol());
}

} // impl

// [[Rcpp::export]]
IntegerVector dims(RObject x) 
{
    RCPP_RETURN_MATRIX(impl::dims, x);
}
```

And checking this against `base::dim`, 

```{r}
classes <- c(
    "integer", "numeric", "raw", "logical",
    "complex", "character", "list", "expression"
)
sapply(seq_along(classes), function(i) {
    x <- matrix(
        vector(mode = classes[i], length = i ^ 2), 
        nrow = i
    )
    all.equal(dims(x), dim(x))
})
```

everything seems to be in order. 

It's worth pointing out that, for various reasons, it is possible to pass a 
matrix object to an Rcpp function which calls `RCPP_RETURN_VECTOR`: 

```{r}
len(1:9)

len(matrix(1:9, 3))
```

Although this is sensible in the case of `len` -- and even saves us from 
implementing a matrix-specific version -- there may be situations where 
this behavior is undesirable. To distinguish between the two object types we 
can rely on the API function `Rf_isMatrix`: 

```{r, engine='Rcpp'}
#include <Rcpp.h>
using namespace Rcpp;

namespace impl {

template <int RTYPE>
int len(const Vector<RTYPE>& x) 
{
    return static_cast<int>(x.size());
}

} // impl

// [[Rcpp::export]]
int len2(RObject x) 
{
    if (Rf_isMatrix(x)) {
        stop("matrix objects not supported.");
    }
    RCPP_RETURN_VECTOR(impl::len, x);
}
```

```{r}
len2(1:9)

tryCatch(
    len2(matrix(1:9, 3)),
    error = function(e) print(e)
)
```

We don't have to worry about the opposite scenario, as this is already handled 
within Rcpp library code: 

```{r}
tryCatch(
    dims(1:5),
    error = function(e) print(e)
)
```

---

### Generic Return Type

In many cases our return type will correspond to our input type. For example, 
exposing the Rcpp sugar function `rev` is trivial: 

```{r, engine='Rcpp'}
#include <Rcpp.h>
using namespace Rcpp;

template <int RTYPE>
Vector<RTYPE> Rev(const Vector<RTYPE>& x)
{
    return rev(x);
}

// [[Rcpp::export]]
RObject rev2(RObject x)
{
    RCPP_RETURN_VECTOR(Rev, x);
}
```

```{r}
rev2(1:5)

rev2(as.list(1:5 + 2i))

rawToChar(rev2(charToRaw("abcde")))
```

As a slightly more complex example, suppose we would like to write a function 
to sort matrices which preserves the dimensions of the input, since 
`base::sort` falls short of the latter stipulation: 

```{r}
sort(matrix(c(1, 3, 5, 7, 9, 2, 4, 6, 8), 3))
```

There are two obstacles we need to overcome:

1. The `Matrix` class does not implement its own `sort` method. However, 
  since [`Matrix` inherits from `Vector`](
  https://github.com/RcppCore/Rcpp/blob/master/inst/include/Rcpp/vector/Matrix.h#L28), 
  we *can* sort the matrix *as a `Vector`* and construct the result from this 
  sorted data with the appropriate dimensions. 
2. As noted previously, the `RCPP_RETURN` macros will generate code to handle 
  exactly 8 `SEXPTYPE`s; no less, no more. Some functions, like `Vector::sort`, 
  are not implemented for all eight of these types, so in order to avoid a 
  compilation error, we need to add template specializations. 

With this in mind, we have the following implementation of `msort`: 

```{r, engine='Rcpp'}
#include <Rcpp.h>
using namespace Rcpp;

// primary template
template <int RTYPE>
Matrix<RTYPE> Msort(const Matrix<RTYPE>& x)
{
    return Matrix<RTYPE>(
        x.nrow(),
        x.ncol(),
        clone(x).sort().begin()
    );
}

// template specializations for raw vectors, 
// lists, and expression vectors
//
// we can just throw an exception, as base::sort 
// does the same
template <>
Matrix<RAWSXP> Msort(const Matrix<RAWSXP>& x)
{ stop("sort not allowed for raw vectors."); }

template <>
Matrix<VECSXP> Msort(const Matrix<VECSXP>& x)
{ stop("sort not allowed for lists."); }

template <>
Matrix<EXPRSXP> Msort(const Matrix<EXPRSXP>& x)
{ stop("sort not allowed for expression vectors."); }

// [[Rcpp::export]]
RObject msort(RObject x)
{
    RCPP_RETURN_MATRIX(Msort, x);
}
```

Note that elements will be sorted in column-major order since we filled our 
result using [this constructor](
https://github.com/RcppCore/Rcpp/blob/master/inst/include/Rcpp/vector/Matrix.h#L63-L69). We can verify that `msort` works as intended by checking a few test cases: 

```{r}
(x <- matrix(c(1, 3, 5, 7, 9, 2, 4, 6, 8), 3))
msort(x)
sort(x)

(x <- matrix(c("a", "c", "z", "y", "b", "x"), 3))
msort(x)
sort(x)

x <- matrix(as.list(1:9), 3); str(x)
tryCatch(
    msort(x),
    error = function(e) print(e)
)
tryCatch(
    sort(x),
    error = function(e) print(e)
)
```

---

### Revisiting the 'ends' Function

Having familiarized ourselves with basic usage of the `RCPP_RETURN` macros, we 
can return to the problem of implementing our `ends` function with 
`RCPP_RETURN_VECTOR`. Just to recap the situation, the template function 
passed to the macro must meet the following two criteria in C++98 mode:

1. It is templated on a single, integer parameter (representing the 
  `Vector` type).
2. It accepts a **single** `SEXP` (or convertible to `SEXP`) **argument**. 

Currently `ends` has the signature 

```{r, engine='Rcpp', eval=FALSE}
template <int RTYPE>
Vector<RTYPE> ends(const Vector<RTYPE>&, int);
```

meaning that the first criterion is met, but the second is not. In order 
preserve the functionality provided by the `int` parameter, we effectively 
need to generate a new template function which has access to the user-provided 
value at run-time, but without passing it as a function parameter. 

The technique we are looking for is called [partial function application](
https://en.wikipedia.org/wiki/Partial_application), and it can be implemented 
using one of my favorite C++ tools: the [functor](
https://en.wikipedia.org/wiki/Function_object). Contrary to typical functor 
usage, however, our implementation features a slight twist: rather than 
using a template class with a non-template function call operator, as is the 
case with [`std::greater`](
http://en.cppreference.com/w/cpp/utility/functional/greater), etc., we are 
going to make `operator()` a template itself: 

```{r, engine='Rcpp'}
#include <Rcpp.h>
using namespace Rcpp;

class Ends {
private:
    int n;

public:
    Ends(int n)
        : n(n)
    {}

    template <int RTYPE>
    Vector<RTYPE> operator()(const Vector<RTYPE>& x)
    {
        n = std::min((R_xlen_t)n, x.size() / 2);
        Vector<RTYPE> res(2 * n);

        std::copy(x.begin(), x.begin() + n, res.begin());
        std::copy(x.end() - n, x.end(), res.begin() + n);

        return res;
    }
};

// [[Rcpp::export]]
RObject ends(RObject x, int n = 6)
{
    RCPP_RETURN_VECTOR(Ends(n), x);
}
```

Not bad, right? All in all, the changes are fairly minor: 

* The function body of `Ends::operator()` is identical to that of 
  `impl::ends`. 
* `n` is now a private data member rather than a function parameter, which 
  gets initialized in the constructor. 
* Instead of passing a free-standing template function to `RCPP_RETURN_VECTOR`, 
  we pass the expression `Ends(n)`, where `n` is supplied at run-time from the 
  R session. In turn, the macro will invoke `Ends::operator()` on the `SEXP` 
  (`RObject`, in our case), using the specified `n` value. 

We can demonstrate this on various test cases: 

```{r}
ends(1:9)

ends(letters, 3)

ends(rnorm(20), 2)
```

---

### A Modern Alternative 

As alluded to earlier, a more modern compiler (supporting C++11 or later) 
will free us from the "single `SEXP` argument" restriction, which means 
that we no longer have to move additional parameters into a function 
object. Here is `ends` re-implemented using the C++11 version of 
`RCPP_RETURN_VECTOR` (note the `// [[Rcpp::plugins(cpp11)]]` 
attribute declaration): 

```{r, engine='Rcpp'}
// [[Rcpp::plugins(cpp11)]]
#include <Rcpp.h>
using namespace Rcpp;

namespace impl {

template <int RTYPE>
Vector<RTYPE> ends(const Vector<RTYPE>& x, int n)
{
    n = std::min((R_xlen_t)n, x.size() / 2);
    Vector<RTYPE> res(2 * n);

    std::copy(x.begin(), x.begin() + n, res.begin());
    std::copy(x.end() - n, x.end(), res.begin() + n);

    return res;
}

} // impl

// [[Rcpp::export]]
RObject ends(RObject x, int n = 6)
{
    RCPP_RETURN_VECTOR(impl::ends, x, n);
}
```

```{r}
ends(1:9)

ends(letters, 3)

ends(rnorm(20), 2)
```

The [current definition](
https://github.com/RcppCore/Rcpp/blob/master/inst/include/Rcpp/macros/dispatch.h#L33-L78) of `RCPP_RETURN_VECTOR` and `RCPP_RETURN_MATRIX` allows for up 
to 24 arguments to be passed; although in principal, the true upper bound 
depends on your compiler's implementation of the `__VA_ARGS__` macro, which 
is likely greater than 24. Having said this, if you find yourself trying 
to pass around more than 3 or 4 parameters at once, it's probably time 
to do some refactoring. 
