---
title: "RcppMLPACK2 and the MLPACK Machine Learning Library"
author: "Dirk Eddelbuettel"
license: GPL (>= 2)
tags: machine_learning armadillo mlpack
summary: "RcppMLPACK2 bring access to MLPACK to R"
---

### mlpack 

[mlpack](http://www.mlpack.org/) is, to quote, _a scalable machine learning library, written in C++,
that aims to provide fast, extensible implementations of cutting-edge machine learning
algorithms_. It has been written by [Ryan Curtin](http://www.ratml.org/) and others, and is
described in two papers in [BigLearning (2011)](http://www.mlpack.org/papers/mlpack2011.pdf) and
[JMLR (2013)](http://www.mlpack.org/papers/mlpack2013.pdf). [mlpack](http://www.mlpack.org/) uses
[Armadillo](http://arma.sf.net) as the underlying linear algebra library, which, thanks to
[RcppArmadillo](http://dirk.eddelbuettel.com/code/rcpp.armadillo.html), is already a rather
well-known library in the R ecosystem.

### RcppMLPACK1

[Qiang Kou](https://github.com/thirdwing) has created the
[RcppMLPACK package on CRAN](https://cran.r-project.org/package=RcppMLPACK) for easy-to-use
integration of [mlpack](http://www.mlpack.org/) with R. It integrates the
[mlpack](http://www.mlpack.org/) sources, and is, as a CRAN package, widely available on all
platforms. 

However, this [RcppMLPACK](https://cran.r-project.org/package=RcppMLPACK) package is also based on a
by-now dated version of [mlpack](http://www.mlpack.org/).  Quoting again: _mlpack provides these
algorithms as simple command-line programs and C++ classes which can then be integrated into
larger-scale machine learning solutions._ Version 2 of the [mlpack](http://www.mlpack.org/) sources
switched to a slightly more encompassing build also requiring the [Boost](https://www.boost.org)
libraries 'program_options', 'unit_test_framework' and 'serialization'.  Within the context of an R
package, we could condition out the first two as R provides both the direct interface (hence no need
to parse command-line options) and also the testing framework.  However, it would be both difficult
and potentially undesirable to condition out the serialization which allows
[mlpack](http://www.mlpack.org/) to store and resume machine learning tasks.

We refer to this version now as [RcppMLPACK1](https://github.com/thirdwing/RcppMLPACK1).

### RcppMLPACK2

As of February 2017, the current version of [mlpack](http://www.mlpack.org/) is 2.1.1.  As it
requires external linking with (some) [Boost](https://www.boost.org) libraries as well as with
[Armadillo](http://arma.sf.net), we have created a new package
[RcppMLPACK2](https://github.com/rcppmlpack/rcppmlpack2) inside a new
[GitHub organization RcppMLPACK](https://github.com/rcppmlpack).

#### Linux 
This package works fine on Linux provided [mlpack](http://www.mlpack.org/),
[Armadillo](http://arma.sf.net) and [Boost](https://www.boost.org) are installed.

#### OS X / macOS

For maxOS / OS X, [James Balamuta](http://thecoatlessprofessor.com/) has tried to set up a homebrew
recipe but there are some tricky interaction with the compiler suites used by both brew and R on
macOS.

#### Windows

For Windows, one could do what [Jeroen Ooms](https://github.com/jeroenooms) has done and build
(external) libraries.  Volunteers are encouraged to get in touch via the issue tickets at GitHub.

#### Installation from source

Release are available from a [drat](http://dirk.eddelbuettel.com/code/drat.html) repository hosted
in the GitHub orgranization RcppMLPACK.  So

```{r, eval=FALSE}
drat:::add("RcppMLPACK")         # first add the repo
install.package("RcppMLPACK2")   # install the pacage
update.packages()                # or update to newer one (if one exists)
```

will use this. If you prefer to rather pick a random commit state, 

```{r, eval=FALSE}
remotes::install_github("rcppmlpack/rcppmlpack2")
```

will work as well.

### Example: Logistic Regression

To illustrate [mlpack](http://www.mlpack.org/) we show a first simple example also included in the
package.  As the rest of the Rcpp Gallery, these are "live" code examples.


```{r, engine="Rcpp"}

#include <RcppMLPACK.h>				// MLPACK, Rcpp and RcppArmadillo

#include <mlpack/methods/logistic_regression/logistic_regression.hpp> 	// particular algorithm used here

// [[Rcpp::depends(RcppMLPACK)]]

// [[Rcpp::export]]
Rcpp::List logisticRegression(const arma::mat& train,
                              const arma::irowvec& labels,
                              const Rcpp::Nullable<Rcpp::NumericMatrix>& test = R_NilValue) {
    
    // MLPACK wants Row<size_t> which is an unsigned representation
    // that R does not have
    arma::Row<size_t> labelsur, resultsur;

    // TODO: check that all values are non-negative
    labelsur = arma::conv_to<arma::Row<size_t>>::from(labels);

    // Initialize with the default arguments.
    // TODO: support more arguments>
    mlpack::regression::LogisticRegression<> lrc(train, labelsur);
    
    arma::vec parameters = lrc.Parameters();

    Rcpp::List return_val;
    
    if (test.isNotNull()) {
        arma::mat test2 = Rcpp::as<arma::mat>(test);
        lrc.Classify(test2, resultsur);
        arma::vec results = arma::conv_to<arma::vec>::from(resultsur);
        return_val = Rcpp::List::create(Rcpp::Named("parameters") = parameters,
                                        Rcpp::Named("results") = results);
    } else {
        return_val = Rcpp::List::create(Rcpp::Named("parameters") = parameters);
    }

    return return_val;

}
```

We can then call this function with the same (trivial) data set as used in the first unit test for
it:

```{r}
logisticRegression(matrix(c(1, 2, 3, 1, 2, 3), nrow=2, byrow=TRUE), c(1L, 1L, 0L))
```


### Example: Naive Bayes Classifier

A second examples shows the `NaiveBayesClassifier` class.


```{r, engine="Rcpp"}
#include <RcppMLPACK.h>				// MLPACK, Rcpp and RcppArmadillo

#include <mlpack/methods/naive_bayes/naive_bayes_classifier.hpp> 	// particular algorithm used here

// [[Rcpp::depends(RcppMLPACK)]]

// [[Rcpp::export]]
Rcpp::List naiveBayesClassifier(const arma::mat& train,
                                const arma::irowvec& labels,
                                const int& classes,
                                const Rcpp::Nullable<Rcpp::NumericMatrix>& test = R_NilValue) {

    // MLPACK wants Row<size_t> which is an unsigned representation
    // that R does not have
    arma::Row<size_t> labelsur, resultsur;

    // TODO: check that all values are non-negative
    labelsur = arma::conv_to<arma::Row<size_t>>::from(labels);

    // Initialize with the default arguments.
    // TODO: support more arguments>
    mlpack::naive_bayes::NaiveBayesClassifier<> nbc(train, labelsur, classes);

    Rcpp::List return_val;
    if (test.isNotNull()) {
        arma::mat armatest = Rcpp::as<arma::mat>(test);
        nbc.Classify(armatest, resultsur);
    
        arma::irowvec results = arma::conv_to<arma::irowvec>::from(resultsur);
        return Rcpp::List::create(Rcpp::Named("means") = nbc.Means(),
                                  Rcpp::Named("variances") = nbc.Variances(),
                                  Rcpp::Named("probabilities") = nbc.Probabilities(),
                                  Rcpp::Named("classification") = results);
    } else {
        return Rcpp::List::create(Rcpp::Named("means") = nbc.Means(),
                                  Rcpp::Named("variances") = nbc.Variances(),
                                  Rcpp::Named("probabilities") = nbc.Probabilities());
    }
}
```

We can use the sample data included in recent-enough version of the RcppMLPACK package:

```{r}
library(RcppMLPACK)
data(trainSet)                ## data part of RcppMLPACK package (when using RcppMLPACK2 source)
trainmat <- t(trainSet[, -5]) ## train data
trainlab <- trainSet[, 5]     ## labels
naiveBayesClassifier(trainmat, trainlab, 2L)                   ## just model
testmat <- t(testSet[, -5])   ## test data
testlab <- testSet[, 5]             
res <- naiveBayesClassifier(trainmat, trainlab, 2L, testmat)   ## also classify
res
## res was a rowvector but comes back as 1-row matrix
all.equal(res[[4]],  testlab)
```


As we can see, the computed classification on the test set corresponds to the expected
classification in `testlabels`.
