---
title: "Using Armadillo with SuperLU"
author: "Qiang Kou and Dirk Eddelbuettel"
license: GPL (>= 2)
tags: armadillo sparse
summary: Using Armadillo with the SuperLU solver
---

[Armadillo](http://arma.sourceforge.net/) is very versatile C++ library for linear algebra, brough
to R via the [RcppArmadillo](http://dirk.eddelbuettel.com/code/rcpp.armadillo.html) package.  It
has proven to be very useful and popular, and is (as of February 2017) used by well over 300 CRAN
packages as indicated by the reverse depends / linking-to on its
[CRAN page](https://cran.r-project.org/package=RcppArmadillo).  Several earlier posts here on the
Rcpp Gallery also demonstrate different usage patterns.

[Armadillo](http://arma.sourceforge.net/) has a core focus on dense matrices, but continues to
expand its capabilities for sparse matrices.  Basic operation are supported directly via the
templated header files, along with calls back into the default (dense) LAPACK and BLAS libraries we
can access easily as R uses them.

[Armadillo](http://arma.sourceforge.net/) also supports dedicated sparse solvers via the
[SuperLU](http://crd-legacy.lbl.gov/~xiaoye/SuperLU/) package.  _However_, this requires access to
the [SuperLU](http://crd-legacy.lbl.gov/~xiaoye/SuperLU/) library.  Many Linux distributions ship
it, see e.g. the [Debian package page](https://packages.debian.org/source/sid/armadillo) and the
[Ubuntu package page](http://packages.ubuntu.com/source/yakkety/superlu); there is also a
[homebrew recipe](https://github.com/Homebrew/homebrew-science/blob/master/superlu.rb) for OS X /
macOS (or other systems using `brew`).  As of this writing, the version in the current Ubuntu
release is behind the version Debian. But it is the 5.2.* version that is in Debian that is also
required with current Armadillo versions 7.700.* so we prepared 'backports' via
[Dirk's PPA repo](https://launchpad.net/~edd/+archive/ubuntu/misc/+packages).)

Recently, a [GitHub issue ticket](https://github.com/RcppCore/RcppArmadillo/issues/120) asked how to
use [SuperLU](http://crd-legacy.lbl.gov/~xiaoye/SuperLU/) along with
[RcppArmadillo](http://dirk.eddelbuettel.com/code/rcpp.armadillo.html).  This post essentially
repeats the main answer, which was spread out over multiple posts, in a single article.

In a nutshell, two things need to happen:

1. One needs to define the required variable `ARMA_USE_SUPERLU` which has to be done before the
Armadillo headers are included.  One possibility (shown below) is a `#define` statement right in the
code file.

2. One needs to tell the linker to use the [SuperLU](http://crd-legacy.lbl.gov/~xiaoye/SuperLU/)
library. This step is of course not perfectly portable, and does require that the library be
installed. (A genuinely portable solution would either need to test for presence of SuperLU, or include its
sources.  Both aspects are beyond the scope of this post._

```{r}
Sys.setenv("PKG_LIBS"="-lsuperlu")
```

Now that R knows about this library, we can present the code requiring it:


```{r, engine="Rcpp"}
// Important: this definition ensures Armadillo enables SuperLU
#define ARMA_USE_SUPERLU 1

#include <RcppArmadillo.h>
// [[Rcpp::depends(RcppArmadillo)]]

using namespace arma;

// [[Rcpp::export]]
void superLuDemo() {
    sp_mat A = sprandu<sp_mat>(1000, 1000, 0.1);

    vec b = randu<vec>(1000);
    mat B = randu<mat>(1000, 5);

    vec x = spsolve(A, b);  // solve one system
    mat X = spsolve(A, B);  // solve several systems

    bool status = spsolve(x, A, b);  // use default solver
    if (status == false)  { Rcpp::Rcout << "no solution" << endl; }

    spsolve(x, A, b, "lapack");   // use LAPACK  solver
    spsolve(x, A, b, "superlu");  // use SuperLU solver

    Rcpp::Rcout << "Done.\n";
}
```

This code snippet defines a function `superLuDemo()` which we can call from R:

```{r}
superLuDemo()
```

As the data generated here is random, we did not bother printing the (dense) result vector of
length 1000.
