---
title: "Documenting Rcpp functions and classes in R packages"
author: "Artem Sokolov"
license: MIT
tags: documentation
date: 2020-01-02
summary: Demonstrates how to add Roxygen blocks in .cpp source
---

## Introduction

[Roxygen2](https://cran.r-project.org/web/packages/roxygen2/index.html) is a convenient way to document functions in R packages. Based on the [Doxygen](https://en.wikipedia.org/wiki/Doxygen) model, it parses relevant bits from the comments and generates the corresponding `man/*.Rd` files. The major strength of this model -- besides not having to write `*.Rd` files by hand -- is that documentation ends up living right next to the functionality it describes, thus enabling easy maintenance and synchronization between the two.

When using Rcpp in package development, the R functions in `RcppExports.R` are generated automatically from the source `.cpp` files. Rcpp also faithfully transcribes comment blocks from `.cpp` to `.R`. We can take advantage of this fact to add documentation directly to the `.cpp` files and have it appear in the package reference manual.

This vignette demonstrates four different ways to document Rcpp functions and classes in R packages. We will create an R package from scratch, add some very basic functionality and show how Roxygen2 blocks appear in the final documentation. Throughout the vignette, we will use several R packages to make our job easier: `devtools`, `usethis` and, of course, `rcpp`. All of these can be installed with the standard `install.packages()` interface:

```{r eval=FALSE}
install.packages( c("devtools", "usethis", "rcpp") )
```

## Create a package

We begin by initializing a package skeleton. The following functions create a new directory and establish a barebones structure for an R package with Roxygen2 and Rcpp support:

```{r eval=FALSE}
devtools::create( "~/test/testpkg" )
devtools::document()
usethis::use_rcpp()
```

Using your favorite text editor, create a new file `~/test/testpkg/R/zzz.R` and add the following lines to it:

```{r eval=FALSE}
#' @useDynLib testpkg
NULL

Rcpp::loadModule("double_cpp", TRUE)
```

The `useDynLib` Roxygen block adds a NAMESPACE entry exposing the functionality of the dynamic library compiled from the .cpp source. Likewise `loadModule` exposes the functionality of an [Rcpp module](http://dirk.eddelbuettel.com/code/rcpp/Rcpp-modules.pdf) that we will use to encapsulate our C++ class. Replace `"double_cpp"` with your desired module name, but note that the name has to match what appears in the `.cpp` file below.

## Add C++ functionality and Roxygen2 comment blocks

All code blocks presented in this section are part of the same contiguous `.cpp` file. For the purposes of this tutorial, assume that the file lives in `~/test/testpkg/src/mult.cpp`. Let's review the four different ways to introduce Roxygen2 comment blocks inside `.cpp` files.

### Approach 1: Standalone functions

The first approach is the most straightforward and mimics traditional usage of Roxygen2 in pure R packages. The comments are placed immediately before a stand-alone function and begin with `//'`, which Rcpp will translate into `#'` in the corresponding `RcppExports.R` file:

```{Rcpp eval=FALSE}
#include <Rcpp.h>

//' Multiplies two doubles
//'
//' @param v1 First value
//' @param v2 Second value
//' @return Product of v1 and v2
//' @export
// [[Rcpp::export]]
double mult( double v1, double v2 ) {return v1*v2;}
```

You may notice two `export` keywords used in the comment block. The first, `@export`, tells Roxygen2 to add a `NAMESPACE` entry that will make the function visibile to the end users of your package. Remove this line if the function should be internal to the package.

The second keyword, `[[Rcpp::export]]`, is the standard Rcpp attribute that makes the function available in R. Since it has no interaction with `Roxygen2`, the comment is annotated with the traditional `//` instead. Remove this line if the function is to remain internal within the C++ code.

### Approach 2: Nested field structure

Classes impose an additional layer of hierarchy by grouping multiple functions and variables. We need to decide how this hierarchy should be represented in the R help pages, which itself follows a flat format. If a class is relatively small, a simple solution is to use the nested field structure to describe individual member functions. The entire class will then appear as a single entry in the final package reference manual.

```{Rcpp eval=FALSE}
//' @name Double
//' @title Encapsulates a double
//' @description Type the name of the class to see its methods
//' @field new Constructor
//' @field mult Multiply by another Double object \itemize{
//' \item Paramter: other - The other Double object
//' \item Returns: product of the values
//' }
//' @export
class Double {
public:
  Double( double v ) : value(v) {}
  double mult( const Double& other ) {return value * other.value;}
private:
  double value;
};
```

### Approach 3: Stand-alone pages for individual class methods

Occasionally, a member function may be complex enough to require its own `?` reference entry. A major advantage of Roxygen2 is that such entries can be created with relative ease by placing the corresponding comment block at the top level in a file.

```{Rcpp eval=FALSE}
//' @name Double$new
//' @title Constructs a new Double object
//' @param v A value to encapsulate
```

The only disadvantage is that the documentation does not live directly next to the function it describes. Unforunately, this is due to a limitation that only the top-level comment blocks are exported into R by Rcpp.

### Approach 4: Rcpp module docstrings

The final place for function documentation is inside the docstring feature provided by the Rcpp modules themselves. This works well for relatively simple classes. Unfortunately, the documentation becomes overly verbose, if a class makes heavy use of templates.

```{Rcpp eval=FALSE}
RCPP_EXPOSED_CLASS(Double)
RCPP_MODULE(double_cpp) {
  using namespace Rcpp;

  class_<Double>("Double")
    .constructor<double>("Wraps a double")
    .method("mult", &Double::mult, "Multiply by another Double object")
    ;
}
```