---
title: Recreating rmultinom and rpois with Rcpp
author: Joshua French
license: GPL (>= 2)
tags: basics
summary: Demonstrates how to mimic the functionality of R's Distributions using Rcpp
---

Sometimes one needs to mimic the exact behavior of R's `Distributions` within C++
code. The incredible Rcpp team has provided access to these distributions through
`Rmath.h` (in the `R::` namespace), as well as through the `Rcpp::` namespace where there
can be two forms: scalar as in R, and vectorized via Rcpp sugar. The behavior of these
functions may not always exactly match what the user expects from the standard R behavior,
particularly if attempting to use the functions in `Rmath.h`. In particular, the functions
in `Rmath.h` are not vectorized. In what follows, I will use Rcpp to mimic the
behavior of both the `rmultinom` and `rpois` functions available in base R so that this
functionality and behavior is provided in native C++.

The [multinomial distribution](https://en.wikipedia.org/wiki/Multinomial_distribution)
generalizes the binomial distribution to `k` discrete outcomes instead of 2; consequently,
it is parameterized in terms of `k` probabilities that must sum to 1. The base R function
`rmultinom` used for generating multinomial data takes three arguments: `n` the number of
simulated data sets to produce, `size`, the number of multinomial outcomes to sample for
each data set, and `prob` a numeric vector of probabilities. The function returns a `k`
$\times$ `n` integer matrix.


The following C++ code uses the `R::rmultinom` function available in `Rmath.h` to generate
`size` multinomial outcomes. The `R::rmultinom` function relies on referencing a pointer
to an `IntegerVector` to store the results. We create a helper function, `rmultinom_1`,
that draws `size` multinomial outcomes from the multinomial distribution based on the
probabilities in `prob`. We then do this `n` independent times in the function
`rmultinom_rcpp`. To match the base R functionality, `rmultinom_rcpp` returns a `k`
$\times$ `n` `IntegerMatrix`.

```{Rcpp chunk1}
#include <Rcpp.h>
using namespace Rcpp;

IntegerVector rmultinom_1(unsigned int &size, NumericVector &probs, unsigned int &N) {
    IntegerVector outcome(N);
    rmultinom(size, probs.begin(), N, outcome.begin());
    return outcome;
}

// [[Rcpp::export]]
IntegerMatrix rmultinom_rcpp(unsigned int &n, unsigned int &size, NumericVector &probs) {
    unsigned int N = probs.length();
    IntegerMatrix sim(N, n);
    for (unsigned int i = 0; i < n; i++) {
        sim(_,i) = rmultinom_1(size, probs, N);
    }
    return sim;
}
```

We now check if the `rmultinom` and `rmultinom_rcpp` functions produce the same results.
We generate a vector of 200 probabilities that sum to 1. We will sample 500 multinomial
outcomes and do this independently 20 times.

```{r chunk2}
prob <- runif(200)
prob <- prob/sum(prob) # standardize the probabilities
size <- 500
n <- 20

set.seed(10)
sim_r <- rmultinom(n, size, prob)
set.seed(10)
sim_rcpp <- rmultinom_rcpp(n, size, prob)
all.equal(sim_r, sim_rcpp)
```

A benchmark of the functions suggests that the `rmultinom_rcpp` function is very slightly
slower than the `rmultinom` function, but that is not really a concern for our purposes.

```{r chunk3}
microbenchmark::microbenchmark(
    rmultinom(1000, size, prob),
    rmultinom_rcpp(1000, size, prob)
)
```

The [poisson distribution](https://en.wikipedia.org/wiki/Poisson_distribution) is a
non-negative discrete distribution characterized by having identical mean and
variance. The base R function `rpois` used for generating Poisson data takes two
arguments: `n` the number of simulated values to produce, and `lambda`, a positive numeric
vector. The `rpois` function cycles (and recycles) through the values in `lambda` for each
successive value simulated. The function produces an integer vector of length `n`. We
provide similar functionality using the `R::rpois` function available in `Rmath.h`. Note
that we cycle through the values of `lambda` so that if the end of the `lambda` vector is
reached before we have generated `n` values, then we restart at the beginning of the
`lambda` vector.


```{Rcpp chunk4}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
IntegerVector rpois_rcpp(unsigned int &n, NumericVector &lambda) {
    unsigned int lambda_i = 0;
    IntegerVector sim(n);
    for (unsigned int i = 0; i < n; i++) {
        sim[i] = R::rpois(lambda[lambda_i]);
        // update lambda_i to match next realized value with correct mean
        lambda_i++;
        // restart lambda_i at 0 if end of lambda reached
        if (lambda_i == lambda.length()) {
            lambda_i = 0;
        }
    }
    return sim;
}
```

We now evaluate whether the `rpois` and `rpois` functions produce the same results. We
generate a positive vector with 200 values for `lambda` and draw `length(lambda) + 5`
independent Poisson values.

```{r chunk5}
lambda <- runif(200, 0.5, 3)
set.seed(10)
pois_sim_r <- rpois(length(lambda) + 5, lambda)
set.seed(10)
pois_sim_rcpp <- rpois_rcpp(length(lambda) + 5, lambda)
all.equal(pois_sim_r, pois_sim_rcpp)
```

A benchmark of the two functions suggests the `rpois_rcpp` function may be slightly faster, 
but once again, that is not our primary concern here.

```{r chunk6}
microbenchmark::microbenchmark(
    rpois(length(lambda) + 5, lambda),
    rpois_rcpp(length(lambda) + 5, lambda)
)
```
