---
title: "Handling R6 objects in C++"
author: "David Quesada"
license: GPL (>= 2)
tags: basics r6 featured
summary: Some examples of how to create R6 objects in C++ and how to call their methods
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo=TRUE)
opts_chunk$set(cache=FALSE)
options(width=80)
```

### Introduction

When we are using R6 objects and want to introduce some C++ code in our 
project, we may also want to interact with these objects using Rcpp. With
this objective in mind, the key to interacting with R6 objects is that
they are essentially environments, and so they can be treated as such.

In this example, we will define a 'Person' class with some private attributes
and public methods. It will allow us to show how to create R6 objects in C++ 
and how to interact with them.

### Creating a dummy R6 class

First, we define the class 'Person', with a name, an id, an item and some 
methods: 

```{r person}
library(R6)

#' R6 class that defines a person
Person <<- R6::R6Class("Person",
  public = list(
    #' @description 
    #' Constructor of the 'Person' class
    #' @param name a string that defines this person's name
    #' @param id an integer that defines this person's id
    #' @return A new 'Person' object
    initialize = function(name, id){
      private$name <- name
      private$id <- id
    },
    
    get_name = function(){return(private$name)},
    
    get_id = function(){return(private$id)},
    
    #' @description 
    #' Gives an item to the person
    #' @param item a string that defines the item
    give_item = function(item){private$item <- item},
    
    #' @description 
    #' A public function that calls a private one
    good_foo = function(){
      return(paste0("Wrapped inside: {", private$bad_foo(), "}"))
    }
  ),
  private = list(
    #' @field name the name of the person
    name = NULL,
    #' @field id the id of the person
    id = NULL,
    #' @field item some item that the person has
    item = NULL,
    
    #' @description 
    #' A private function that should not be called from outside the object
    bad_foo = function(){return("This is a private function")}
  )
)
```

With this simple class, we will be able to create R6 objects, initialize them 
and call some of their methods.

### Creating R6 objects in C++

To create R6 objects, we must first get the 'new' function that initializes
them. In this case, given that we have loaded the Person class into the 
global environment, we will get it from there:

```{r init, engine="Rcpp"}
#include<Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
List initialize_list(StringVector &names, IntegerVector &ids){
  List res(names.size());
  Environment g_env = Environment::global_env();
  Environment p_env = g_env["Person"];
  Function new_person = p_env["new"];
  
  for(unsigned int i = 0; i < names.size(); i++){
    Environment new_p;
    String name = names[i];
    int id = ids[i];
    new_p = new_person(name, id);
    res[i] = new_p;
  }
  
  return res;
}

```

```{r initialize_list, eval=TRUE}
names <- c("Jake", "Anne", "Alex")
ids <- 1:3

res <- initialize_list(names, ids)

print(res)
```

The previous example initializes a list with Persons. It is important to notice 
that the only function relevant to us from the global environment is the 'new' 
function created by the R6 class. All the functions defined inside Person will
be contained in each instance of the class.

If the R6 object is defined inside some package, included our own package if 
we are developing one, we must get it from there. To do this, the previous 
function has to be modified as follows:

```{r modified_init, engine="Rcpp", eval=TRUE}
#include<Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
List initialize_list(unsigned int size){
  List res(size);
  Environment package_env("package:some_package");
  Environment class_env = package_env["some_class"];
  Function new_instance = class_env["new"];
  
  for(unsigned int i = 0; i < size; i++){
    Environment new_i;
    new_i = new_instance();
    res[i] = new_i;
  }
  
  return res;
}

```

### Calling the methods of an instance

Once we have instantiated an R6 object as an environment, we can call its
methods after getting them from the instance:

```{r method item, engine="Rcpp"}
#include<Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
Environment create_person_item(){
  Environment g_env = Environment::global_env();
  Environment p_env = g_env["Person"];
  Function new_person = p_env["new"];
  Environment new_p;
  String name = "Jake";
  int id = 1;
  String item = "shovel";
  
  new_p = new_person(name, id);
  Function give_i = new_p["give_item"];
  give_i(item);

  return new_p;
}

```

```{r init_w_item, eval=TRUE}
res <- create_person_item()

print(res)
```

Note that if we are creating multiple instances, we have to get the desired
method from each one of them.

Private attributes and methods cannot be accessed in this manner, trying to do 
so results in an error:

```{r method private, engine="Rcpp"}
#include<Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
String access_method(std::string foo_name){
  Environment g_env = Environment::global_env();
  Environment p_env = g_env["Person"];
  Function new_person = p_env["new"];
  Environment new_p;
  String name = "Jake";
  int id = 1;
  
  new_p = new_person(name, id);
  Function foo = new_p[foo_name];
  String res = foo();
  
  return res;
}

```

```{r access private, eval=TRUE}
tryCatch({res <- access_method("bad_foo")},
         error=function(cond){print(paste0("Exception raised: ", cond))})

res <- access_method("good_foo")

print(res)
```

The error is telling us that there is no function called "bad_foo" in
the environment of the object "new_p". In that environment, only public 
attributes and methods are present on the surface. Theoretically, we could
still try to find the private attributes and methods inside the object, but
it goes against the paradigm and it would be much easier to just make 
the method we want to call public.

### Passing down R6 objects

Passing R6 objects as arguments to C++ functions is also straight forward
by treating them as environments:

```{r pass down R6, engine="Rcpp"}
#include<Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
void give_shovel(Environment &person){
  String item = "shovel";

  Function give_i = person["give_item"];
  give_i(item);
}

```

```{r pass R6 to C++}
p <- Person$new("Jake", 1)

give_shovel(p)

print(p)
```

Note that there is no need to return the object, as the environment is passed
by reference and changes inside it will be reflected outside the function.
