---
title: "Mixing Rcpp modules and Rcpp attributes"
author: "Bob Jansen"
license: GPL (>= 2)
tags: basics modules
summary: How to use Rcpp modules in combination with Rcpp attributes
---

### Introduction

With Rcpp attributes Rcpp modules (described in the Rcpp vignettes) it is
easy to expose \proglang{C++} classes and functions to \proglang{R}. This note
describes how to use classes exported by modules in conjunction with functions
exported using Rcpp attributes through the use of \code{RCPP_EXPOSED\*} macros.

In the following snippets, a simple example is given of a simple class and a
function that takes that class as an argument. The \proglang{C++} function is
exported using Rcpp attributes as follows:

```{r chunk-1, engine="Rcpp"}
#include <Rcpp.h>

// [[Rcpp::export]]
void shout(std::string message) {
    Rcpp::Rcout << message << "!" << std::endl;
}
```
Now, calling this function from \code{R} is as easy as one can hope:
```{r}
shout("Hello World")
```

\proglang{C++} classes can be exported using Rcpp modules. The simple class
\code{Echo} below has a \code{get()} method which returns the original
constructor parameter.

```{r chunk-2, engine="Rcpp", eval=FALSE}
#include <Rcpp.h>
#include <string>

using namespace Rcpp;

class Echo {
private:
      std::string message;
public:
      Echo(std::string message) : message(message) {}

      std::string get() {
        return message;
      }
};
```

This class can now be exposed to \proglang{R} by specifing the constructors and
the methods that should be callable from \proglang{R} with

```{r chunk-3, engine="Rcpp", eval=FALSE}
RCPP_MODULE(echo_module) {
      class_<Echo>("Echo")
      .constructor<std::string>()
      .method("get", &Echo::get)
      ;
};
```

Unfortunately, combining these two snippets as above  creates a problem. The Rcpp
attributes machinery that exports \code{shout()} will not be automagically
aware of the \code{Echo} class. This will cause an error when the package is
loaded by \proglang{R} as the required functionality that transforms the class
between a \code{SEXP} and a regular \proglang{C++} object can't be loaded. The
solution is simple: instruct the compiler to do so explicitly using the
\code{RCPP_EXPOSED\*} family of macros.  In the current case it suffices to add
```{r chunk-4, engine="Rcpp", eval=FALSE}
RCPP_EXPOSED_AS(Echo)
```

```{r ref.label=c(paste0('chunk-', 1:4)), engine="Rcpp", echo=FALSE}
```

Now, constructing and using the class from \proglang{R} is again
straightforward

```{r}
echo <- new(Echo, "Hello World")
echo$get()
shout(echo$get())
```

### The \code{RCPP_EXPOSED\*} macros

Rcpp defines a number \code{RCPP_EXPOSED\*}macros in
\code{inst/include/Rcpp/macros/module.h}, the most important ones are

- \code{RCPP_EXPOSED_AS}  which allows passing objects from \proglang{R} to
  \proglang{C++}. As seen above, this is needed when exported functions want to
  take a \proglang{C++} object as argument. Other uses include methods and
  constructors of other Rcpp modules classes that take a \proglang{C++} object
  as argument;
- \code{RCPP_EXPOSED_WRAP} which allows the other way around; This is needed
  when a exported function or method wants to return a \proglang{C++} object;
- \code{RCPP_EXPOSED_CLASS} which allows both.

